\chapter{Introduction}

\section{Project Context and Objectives}

Context of this project is ongoing research for high performance network interface cards at the Integrated Electronic Systems Labs by Boris Traskov, also supervisor of this project.

In an environment with an ever increasing amount of created data by all kinds of devices, the amount of transferred data over networks (i.e. the internet) increases dramatically. This generates a demand for high speed network interfaces, on the one hand and high performance, low cost full stack network implementations on the other hand. 

Service providers and back-end nodes need to be able to transfer large amounts of data to multiple receivers concurrently, requiring high speed network interfaces. To circumvent -- potentially unnecessary -- \gls{cpu} utilization, much of the work could be "offloaded" to a network interface, providing a higher level of abstraction to the transfer of data, than current state of the art systems provide.

In a future scenario with all devices being connected to each other, often described as the "internet of things", an increasing demand for simple to implement network interfaces, not requiring the presence of an operating system or high performance processors, will be created.

Objectives for the whole project entitled "Design of an Accelerated Event-based Server" is to setup a hardware system on a \textit{Xilinx} \gls{fpga} utilizing \gls{ip} cores, running \textit{nginx} (an event-based web server) on top. Furthermore benchmarks have to be conducted, to prove the strength of such a system. The ideal outcome of this project would be to prove that the system is capable of utilizing the full bandwidth of a Gigabit Ethernet interface. Therefore break downs of the server system on attempted \gls{dos} attacks would belong to the past.

The first part of this project included the design of a System on Chip, as well as getting an operating system up and running on top of it. This was implemented in a preceding project seminar, May to September 2012. Reached accomplishments of this project seminar are documented in the corresponding report \cite{projectseminar}. All results are also briefly summarized in chapter \ref{cha:preceding} of this thesis.

\section{Hardware Platform and Tool Set}

All development, measurements and tests were done on the \textit{Xilinx XUPV5-LX110T} evaluation board.

The \textit{Xilinx XUPV5-LX110T} evaluation board is a modified version of the \textit{ML505} board for universities \cite{xupv5manual}. The difference between the two boards is a larger \gls{fpga} chip on the \textit{XUPV5-LX110T}, containing the \textit{Virtex-5 XC5VLX110T} with 17,280 slices\footnote{Basic logic unit of an \gls{fpga} \cite{fpga_ni}} and four in hardware implemented Ethernet \gls{mac} cores, whereas the \textit{ML505} board contains the \textit{XC5VLX50} with only 7,200 slices and no Ethernet \gls{mac} cores implemented in hardware.

Building the hardware design for the \gls{fpga} was started using the \textit{\gls{xps}} and continued with the \textit{Xilinx ISE Project Navigator}, for a finer control over the parameters of used tools and better logging and reporting capabilities. All used \textit{Xilinx} tools are part of the \textit{Xilinx ISE Design Suite} in version 14.1. The synthesis tool (xst) of this version of the tool suite features parallel synthesis speeding up the development process heavily.

Initial software development for testing purpose was done using the \textit{Xilinx Software Development Kit (SDK)}. The \textit{Linux kernel} itself and all software build on top was configured and compiled on a virtual machine running \textit{Ubuntu Linux} in version 12.04.